Interactive Curvature Tensor Visualization on Digital Surfaces\thanks{This work has been mainly funded by \textsc{DigitalSnow} ANR-11-BS02-009, \textsc{KIDICO} ANR-2010-BLAN-0205 and \textsc{PRIMES Labex} ANR-11-LABX-0063/ ANR-11-IDEX-0007 research grants.}

### 2015-10-30 latex/article.tex

@@

 \institute{ Universit\'e de Lyon, CNRS\\
   LIRIS, UMR5205, F-69621, France
%   \email{\{david.coeurjolly,jeremy.levallois\}@liris.cnrs.fr}
   \and
Universit\'e de Savoie, CNRS\\
LAMA, UMR5127, F-73776, France\\
%\email{jacques-olivier.lachaud@univ-savoie.fr}
}

@@

\begin{abstract}\sloppy
  Interactive visualization is a very convenient tool to explore
  complex scientific data or to explore different parameter settings
  for a given processing algorithm. In this article, we present a tool
  to efficiently explore the curvature tensor on the boundary of
  potentially large and dynamic digital objects (mean and gaussian curvatures,
  principal curvatures, principal directions and normal vector
  field). More precisely, we combine a fully parallel pipeline on GPU
  to extract an adaptive triangulated isosurface of the digital
  object, with a curvature tensor estimation at each surface point
  based on integral invariants. Integral invariants being parametrized
  by a given ball radius, our proposal allows us to explore
  interactively different radii and thus select the appropriate scale
  \Jaco{at} which the computation is performed and visualized.

@@

Volumetric objects are being more and more popular in many
applications ranging from object modeling and rendering in Computer
Graphics to geometry processing in Medical Imaging or Material
Sciences. When considering large volumetric data, interactive
%visualization of such images (or isovalue surfaces) is a complex
visualization of those objects (or isosurfaces) is a complex
problem. Such issues become even more difficult when dynamic
%volumetric datasets are considered. Beside such isosurface
volumetric datasets are considered. Beside
visualization, we are also interested in performing geometry
processing on the digital object and to explore different parameter
settings of the geometry processing tool.  In this work, we focus on
interactive visualization of volumetric data isosurfaces combined with
a curvature tensor (mean/Gaussian curvature, principal curvatures
directions, \ldots) estimation at each point of the digital
surface.
%Curvature maps on a given isosurface are key information when
%exploring 3D structures in Material Sciences \todo{REF}.
Furthermore,
most curvature estimators require a parameter fixing the scale
\Jaco{at} which the computation is performed. For short, such parameter
(integration radius, convolution kernel size, \ldots) specifies \Jaco{a scale for analyzing the object surface, and is naturally related to the amount of perturbation on input data}. As a consequence, exploring interactively
different values of such parameter is mandatory.
\vspace{0.2cm}

@@

\sloppy\noindent\textbf{Related works}\quad Extracting and visualizing
isosurface on volumetric data has been widely investigated since the
seminal Marching Cubes approach by \textsc{Lorensen} and
\textsc{Cline}~\cite{lorensen1987marching}. This approach being data parallel,
GPU implementation of this method is very efficient
\cite{tatarchuk2007real}. However, such technique generates a lot of
triangles which is not well suited to large digital data. Hence,
adaptive approaches have been proposed in order to optimize the
\Jaco{triangulated} mesh resolution \Jaco{according to the object geometry or the
viewpoint}. In this topic, many solutions have been developed in
Computer
Graphics~\cite{shu1995adaptive,schaefer2004dual,lengyel2010voxel,DBLP:journals/cgf/LewinerMPPL10,DBLP:journals/cvgip/LobelloDD14}.
The method developed by \textsc{Lengyel} \textit{et al.}
\cite{lengyel2010voxel} \Jaco{suits best our needs}.
%However, it
%was only developed on CPU to visualize in real-time static or near
%static data, combining octree with new Marching Cubes configurations
%to generate adaptive meshes.
It combines an octree space partitioning with new Marching Cubes configurations
to generate adaptive meshes on CPU from
%, but was only developed on the CPU to visualize
static or near static data.
We propose here a full GPU pipeline
%inspired from this method.
inspired \Jaco{by} this method,
%using a GPU-friendly Lod criterion
\Jaco{that maintains} a view dependent triangulation.
%It leaves the CPU free for other tasks and avoids synchronization with the GPU.
%Our high framerate also allows
%Our high framerate allows this method to run in real-time on fully dynamic data.
Our high framerate allows \Jaco{us} to inspect dynamic 3D data in real-time.
%The triangulation is controlled by a GPU LoD criterion
%that ensures that
%triangles project onto more than one pixel, thus removing geometrical
%aliasing.

Curvature estimation on discrete or digital surface has also been
widely investigated. In \cite{CVIU2014}, authors propose digital
versions of Integral Invariant estimators
\cite{Pottmann2007,Pottmann2009} in order to estimate the complete
curvature tensor (mean/Gaussian curvatures, principal curvatures,
principal curvature directions, normal vector field) on digital
surfaces. Such approaches are based on an integration principle using
a \Jaco{ball} kernel of a given radius. Additionally, authors have
demonstrated that these estimators have multigrid convergence
properties. In this article, we also propose an efficient GPU
implementation of such estimators to visualize
%, in real-time,
such curvature \Jaco{fields} in real-time and
%allowing us
to interactively change the
%explore different
value of the kernel radius.

@@

In \cite{DGCI2013}, authors have defined the 2D digital curvature estimator
$\CurvH{R}$ and the 3D digital mean curvature estimator $\MeanCurvH{R}$ based on
the digital volumetric integral $\VolC(Y,h) \EqDef h^d \MCard(Y)$ \Jaco{(we denote such
quantity $\AreaC(Y,h)$ in dimension 2):}
%% Redite
%% \Jaco{Given some point $\vx \in \R^d$, the set $Y$ is the digital intersection of $\DigShape$ with the ball of radius $R/h$ and center $\vx/h$.}
%% where $Y$ is the intersection of $\DigShape$ with a ball
%% $\Ball{R/h}{\vx/h}$ with $\MCard(Y)$, the number of digital points of
%% the intersection between a ball $\Ball{R/h}{\vx/h}$ of digital radius
%% $R/h$ and $\DigShape$ (we denote such quantity $\AreaC(Y,h)$ in
%% dimension 2):
%
\begin{Definition}
  %
  Given the Gauss digitization $Z\EqDef\DSh$ of a shape $\Shape \subset \R^2$
  (or $\R^3$ for the 3D mean curvature estimator), {\em digital curvature
  estimators} are defined for any point $\vx \in \R^2$ (or $\R^3$) as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    &\CurvH{R}(Z,\vx,h) \EqDef \frac{3\pi}{2R}
    - \frac{3\AreaC(\Ball{R/h}{\vx/h} \cap
    {Z}, h)}{R^3}\,,\label{eq:dig-curvature-estimator-k}\\
    &\MeanCurvH{R}(Z,\vx,h) \EqDef \frac{8}{3R}
    - \frac{4\VolC(\Ball{R/h}{\vx/h} \cap
    Z, h)}{\pi R^4}\,.\label{eq:dig-curvature-estimator-H}
  \end{align}
  %
  where $\Ball{R/h}{\vx/h}$ is the ball of digital radius $R/h$ centered on
  digital point $(\vx/h) \in Z$.
  %
  % \AreaC(\Ball{R/h}{\vx/h} \cap \Body{\DSh}{h}, h) \EqDef h^2 \MCard(
  % \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ and $\VolC(\Ball{R/h}{\vx/h} \cap
  % \Body{\DSh}{h}, h) \EqDef h^3 \MCard( \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ are
  % the number of digital points in the intersection between $\Body{\DSh}{h}$ and
  % the ball $\Ball{R/h}{\vx/h}$ of radius $R/h$ centered on $\vx/h$, scaled by
  % $h^d$.\todo[inline]{J: Hum... pas top...}
  %
  % Similarly, $\hat{\lambda}_1$
  % and $\hat{\lambda}_2$ are the two greatest eigenvalues of the covariance
  % matrix of the digital points in the intersection between $Z$ and the ball.
  %
\end{Definition}
%
They have proved that estimating both 2D curvature and 3D mean curvature on digital
objets with volumetric integral information have the multigrid convergence
property \cite{DGCI2013}. For short, when the digital objet becomes finer and
finer, \ie when the digitization step $h$ tends to zero, the estimated quantity
on $\Bd{\Body{\DSh}{h}}$ converges (theoretically and experimentally) to the
associated one on $\Bd{\Shape}$.
%
%Since our digital estimators are defined on
%digital shapes, in order to compare results with the exact curvature on
%Euclidean shape, we need to match points on boundary of the digital
%shape and points on the border of the Euclidean one.
%
Setting $R = kh^\frac{1}{3}$ on convex shapes
with at least $C^3$-boundary and bounded curvature, where $k$ is a constant
related to the maximal curvature of the shape, they have:
%
\begin{align}
\forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
\forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
&\CurvH{R}(\DSh, \vxH, h) = \Curv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,,\\
&\MeanCurvH{R}(\DSh, \vxH, h) = \MeanCurv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,.
\end{align}

@@

Additionally, eigenvectors associated to $\hat{\lambda}_1$ and
$\hat{\lambda}_2$ of the covariance matrix \Jaco{are} principal
curvature direction \Jaco{estimators} $\PrincDirH{1}{R}$ and
$\PrincDirH{2}{R}$. The smallest eigenvector corresponds to the normal
direction $\NormalDirH{R}$ at $\vx$. Convergence results can be found
in \cite{ChapterIICurvature}.
%%% Changer avec le chapitre de Jaco à un moment

@@

Representing a hierachical structure on GPU is usually challenging since
such a data parallel component is unable to handle recursivity.

@@

Similarly to original MC algorithm, this approach is well suited %perfectly adapted
to a GPU implemntation: given a set of cells (a vector of morton codes),
each triangle patch can be extracted in parallel for both regular cells and transition cells.
%(done by the \emph{geometry shader} in the GPU graphic pipeline).

@@

As illustrated in Figure~\ref{fig_quadtree_partitionning}-\emph{middle}, we
%propose a viewpoint dependent criterion to decide if a cell needs to be refined:
propose a viewpoint dependent criterion to decide if a cell needs to be refined:
the closer we are to the camera, the finer the cells are.
%In addition to the
%distance criterion, we define an angular criterion to prevent refining cells
%outside the camera frustrum.
Such a criterion is also well suited to GPU since it can be evaluated
independently on \Jaco{every} cell.
%Hence, we can efficiently determine if a cell must be subdivided or not.
Figure~\ref{fig_lod_octree} illustrates our level of
details (LoD for short) criterion.
% we use. More precisely,
In dimension 2, if $\alpha$ denotes the viewing angle, an object at a distance $d$ from the camera
%with diameter greater than $2\cdot d\cdot\tan(\alpha)$ is projected into more
has a projected size on screen of $2\cdot d\cdot\tan(\alpha)$.
%than one pixel
(see Figure~\ref{fig_lod_octree}-\emph{left}).
%Hence, the
Our distance criterion is based on the ratio (\emph{visibility ratio} in the
following) between the cell diameter $l(c)$ (power of 2 depending on the depth),
and its projected size. %$2\cdot d_c\cdot\tan(\alpha)$, $d_c$ being the distance between the cell
%center and the camera.
For a given cell $c$, split and merge decision are based
on this visibility ratio:
%
\begin{itemize}
  \item $c$ is split if its children
cells have a visibility ratio greater than  constant $k$\,;
  \item $c$ and its sibling cells are merged if their parent cell $c'$ has a visibility ratio lower than $k$\;
  \item otherwise, the cell $c$ stays for the next frame.
\end{itemize}
%
\begin{figure}[!h]
  \begin{center}
    \begin{overpic}[width=3.5cm]{figs/criterion}
      \put(33,40){$\alpha$}
      \put(50,45){$d_c$}
      \put(95,65){$l(c)$}
    \end{overpic}~~~~~
    \includegraphics[width=3cm]{viewlod2_small}~~~~~
    %\includegraphics[width=3cm]{subdivision}
  \end{center}
  \caption{Notations \emph{(left)} and adaptive meshing in dimension 2 using the LoD distance and angular criterion \emph{(right)}.
  %Based on LoD criterion evaluation, the cell buffer is updated between each frame \emph{(right)}.
  }
  \label{fig_lod_octree}
\end{figure}
%
%In addition to the distance criterion, we add an angular term in order
%to prevent subdivision of cells far from the camera frustrum (see
%Figure~\ref{fig_lod_octree}-\emph{middle}). We do not go into the
%details but please note that stay, split and merge decisions of all cells are computed in
%parallel from all cell morton codes.
Using such a criterion, split and merge decisions are computed in
%of all cells are computed in
parallel from the morton codes of all cells.

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{figs/pipeline2}
  \caption{\Jaco{GPU pipeline summary}. Data buffers are
    represented in green and computations in red. Each computation
    retrieves data from a buffer and fills a new one. }
  \label{fig_pipeline}
\end{figure}

@@

The general idea is to perform an Integral Invariant computation on
GPU at each vertex of the generated \Jaco{triangulated} mesh. First note that
when the LoD criterion is removed, each MC vertex is exactly centered at
a surfel center. At different depth of the octree, MC vertices still
correspond to surfel centers of subsampled versions of the input
object. Hence, Integral Invariant framework defined in
Section~\ref{sec:curv-tens-estim} is consistent with the \Jaco{triangulated}
surface obtained on GPU: triangles are used for visualization purposes
but all computations are performed on the digital object $\DSh$ \Jaco{for}
estimators
(\ref{eq:dig-curvature-estimator-k}),
(\ref{eq:dig-curvature-estimator-H}), (\ref{eq:dig-curvature-estimator-k1})
and (\ref{eq:dig-curvature-estimator-k2}).

To implement the integration on $\Ball{R}{\vx} \cap \Shape$
(Fig.~\ref{fig:approx}-$a$), a naive approach consists in scanning all digital
points inside such domain (Fig.~\ref{fig:approx}-$b$) and then to estimate the
geometrical moment as the sum of the geometrical moments of each elementary
cubes lying in the intersection. On GPU, \emph{mipmap} textures are used to
obtain a multi-resolution information: if the input binary object is stored in a
3D texture, GPU hardware can construct a multi-resolution pyramid such that 8
neighboring voxel intensities at level $l$ are averaged to define the voxel
value at level $l+1$. If the level 0 corresponds to the binary input object, at
a given level $l$, a voxel $(x,y,z)$ contains the fraction of $\DSh$ belonging
to the cube of center $(x,y,z)$ and edge length $2^l$. As a consequence, we can
approximate the volume of $\Ball{R}{\vx}\cap\Shape$ by considering mipmap values
at a given resolution $l$ (Fig.~\ref{fig:approx}-$c$). In this case, errors only
occurs for cells lying inside $\Shape$ (with density 1) not entirely covered by
$\Ball{R}{\vx}$. Furthermore, the \emph{mipmap} texture can be used to design,
using a single texture probe, a fast inclusion test of a given cell $c$ at level
$l$ into the shapes: we say that $c$ is in $\Shape$ if its density (retrieved
from the texture probe at level l) is greater than $1/2$. The idea here is to
mimic a \Jaco{blend ?} of adaptive Gauss digitization process.

@@

Using the hierachical nature of the mipmap texture, we could also
consider hierarchical decompositions of either $\Ball{R}{\vx}$
(Fig.~\ref{fig:approx2}-$a$) or $\Ball{R}{\vx}\cap\Shape$
(Fig.~\ref{fig:approx2}-$b$). The idea would be to decompose the
regions into mipmap cells of different resolution in order to limit
the number of texture access (important bottleneck on GPU hardwares)
and to get better approximated quantities. Note that the scenario in
Fig.~\ref{fig:approx2}-$b$ is clearly the best approach: the
volume is exact  (\wrt the regular grid case) since geometrical moments
are additive quantities, and the number of texture fetches is  optimal
at this point.

@@

\paragraph{Hierarchical Decomposition.}  As discussed above, best
results in terms of texture fetch can be obtained by the approach
presented in Fig.~\ref{fig:approx2}-$b$. At each vertex, the
computation algorithm would consist in a recursive decomposition of
the domain such that a cell is subdivided if it does not belong to
$\Ball{R}{\vx}\cap\Shape$. However, on the GPU %hardware,
this approach is
highly inefficient since the hardware optimizes the parallelism only
when the micro-program described in the shader has a predictive
execution flow. Hence, implementing the recursive algorithm (or its
de-recursified version) requires a lot of branches (conditional
\emph{if} tests) in the flow. Instead of implementing this approach,
we have considered a fast hierarchical approximation following
Fig.~\ref{fig:approx2}-$a$: for a given radius $R$, we precompute a
hierachical decomposition of $B_R$. Such decomposition is made of
octree cells at different resolutions. At a given MC vertex $\vx$, the
algorithm becomes simple since we just scan the ball cells and
estimate the area (or other moments) from the mipmap values associated
to this cell. Note that cells in the $B_R$ decomposition may not be
aligned with mipmap cells. However, we can use GPU %mipmap
functionalities which interpolates mipmap values if we probe at
non-discrete positions. For a given radius $R$, the expected number of
cells \Jaco{in} $B_R$ is in $O(\log{R})$. Implementation details on the
hierarchical octree representation can be found in the supplementary
material (see Section \ref{sec:discussion}).

>>>>

% Copyright 2015
% Hélène Perrier <helene.perrier@liris.cnrs.fr>
% Jérémy Levallois <jeremy.levallois@liris.cnrs.fr>
% David Coeurjolly <david.coeurjolly@liris.cnrs.fr>
% Jean-Philippe Farrugia <jean-philippe.farrugia@liris.cnrs.fr>
% Jean-Claude Iehl <jean-claude.iehl@liris.cnrs.fr>
% Jacques-Olivier Lachaud <jacques-olivier.lachaud@univ-savoie.fr>
%
% This file and related figures are under Creative Commons CC BY-NC-SA 2.0
% See <https://creativecommons.org/licenses/by-nc-sa/2.0/>

@@

 \institute{ Universit\'e de Lyon, CNRS\\
   LIRIS, UMR5205, F-69621, France
%   \email{\{david.coeurjolly,jeremy.levallois\}@liris.cnrs.fr}
   \and
Universit\'e de Savoie Mont Blanc, CNRS\\
LAMA, UMR5127, F-73776, France\\
%\email{jacques-olivier.lachaud@univ-savoie.fr}
}

@@

\begin{abstract}\sloppy
  Interactive visualization is a very convenient tool to explore
  complex scientific data or to explore different parameter settings
  for a given processing algorithm. In this article, we present a tool
  to efficiently explore the curvature tensor on the boundary of
  potentially large and dynamic digital objects (mean and Gaussian curvatures,
  principal curvatures, principal directions and normal vector
  field). More precisely, we combine a fully parallel pipeline on GPU
  to extract an adaptive triangulated isosurface of the digital
  object, with a curvature tensor estimation at each surface point
  based on integral invariants. Integral invariants being parametrized
  by a given ball radius, our proposal allows us to explore
  interactively different radii and thus select the appropriate scale
  at which the computation is performed and visualized.

@@

Volumetric objects are being more and more popular in many
applications ranging from object modeling and rendering in Computer
Graphics to geometry processing in Medical Imaging or Material
Sciences. When considering large volumetric data, interactive
%visualization of such images (or isovalue surfaces) is a complex
visualization of those objects (or isosurfaces) is a complex
problem. Such issues become even more difficult when dynamic
%volumetric datasets are considered. Beside such isosurface
volumetric datasets are considered. Beside
visualization, we are also interested in performing geometry
processing on the digital object and to explore different parameter
settings of the geometry processing tool.  In this work, we focus on
interactive visualization of volumetric data isosurfaces combined with
a curvature tensor (mean/Gaussian curvature, principal curvatures
directions, \ldots) estimation at each point of the digital
surface.
%Curvature maps on a given isosurface are key information when
%exploring 3D structures in Material Sciences \todo{REF}.
Furthermore,
most curvature estimators require a parameter fixing the scale
at which the computation is performed. For short, such parameter
(integration radius, convolution kernel size, \ldots) specifies a scale for analyzing the object surface, and is naturally related to the amount of perturbation on input data. As a consequence, exploring interactively
different values of such parameter is mandatory.
\vspace{0.2cm}

@@

\sloppy\noindent\textbf{Related works}\quad Extracting and visualizing
isosurface on volumetric data has been widely investigated since the
seminal Marching Cubes approach by \textsc{Lorensen} and
\textsc{Cline}~\cite{lorensen1987marching}. This approach being data parallel,
GPU implementation of this method is very efficient
\cite{tatarchuk2007real}. However, such technique generates a lot of
triangles which is not well suited to large digital data. Hence,
adaptive approaches have been proposed in order to optimize the
triangulated mesh resolution according to the object geometry or the
viewpoint. In this topic, many solutions have been developed in
Computer
Graphics~\cite{shu1995adaptive,schaefer2004dual,lengyel2010voxel,DBLP:journals/cgf/LewinerMPPL10,DBLP:journals/cvgip/LobelloDD14}.
The method developed by \textsc{Lengyel} \textit{et al.}
\cite{lengyel2010voxel} suits best our needs.
%However, it
%was only developed on CPU to visualize in real-time static or near
%static data, combining octree with new Marching Cubes configurations
%to generate adaptive meshes.
It combines an octree space partitioning with new Marching Cubes configurations
to generate adaptive meshes on CPU from
%, but was only developed on the CPU to visualize
static or near static data.
We propose here a full GPU pipeline
%inspired from this method.
inspired by this method,
%using a GPU-friendly Lod criterion
that maintains a view dependent triangulation.
%It leaves the CPU free for other tasks and avoids synchronization with the GPU.
%Our high framerate also allows
%Our high framerate allows this method to run in real-time on fully dynamic data.
Our high framerate allows us to inspect dynamic 3D data in real-time.
%The triangulation is controlled by a GPU LoD criterion
%that ensures that
%triangles project onto more than one pixel, thus removing geometrical
%aliasing.

Curvature estimation on discrete or digital surface has also been
widely investigated. In \cite{CVIU2014}, authors propose digital
versions of Integral Invariant estimators
\cite{Pottmann2007,Pottmann2009} in order to estimate the complete
curvature tensor (mean/Gaussian curvatures, principal curvatures,
principal curvature directions, normal vector field) on digital
surfaces. Such approaches are based on an integration principle using
a ball kernel of a given radius. Additionally, authors have
demonstrated that these estimators have multigrid convergence
properties. In this article, we also propose an efficient GPU
implementation of such estimators to visualize
%, in real-time,
such curvature fields in real-time and
%allowing us
to interactively change the
%explore different
value of the kernel radius.

@@

In \cite{DGCI2013}, authors have defined the 2D digital curvature estimator
$\CurvH{R}$ and the 3D digital mean curvature estimator $\MeanCurvH{R}$ based on
the digital volumetric integral $\VolC(Y,h) \EqDef h^d \MCard(Y)$ (we denote such
quantity $\AreaC(Y,h)$ in dimension 2):
%% Redite
%% \Jaco{Given some point $\vx \in \R^d$, the set $Y$ is the digital intersection of $\DigShape$ with the ball of radius $R/h$ and center $\vx/h$.}
%% where $Y$ is the intersection of $\DigShape$ with a ball
%% $\Ball{R/h}{\vx/h}$ with $\MCard(Y)$, the number of digital points of
%% the intersection between a ball $\Ball{R/h}{\vx/h}$ of digital radius
%% $R/h$ and $\DigShape$ (we denote such quantity $\AreaC(Y,h)$ in
%% dimension 2):
%
\begin{Definition}
  %
  Given the Gauss digitization $Z\EqDef\DSh$ of a shape $\Shape \subset \R^2$
  (or $\R^3$ for the 3D mean curvature estimator), {\em digital curvature
  estimators} are defined for any point $\vx \in \R^2$ (or $\R^3$) as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    &\CurvH{R}(Z,\vx,h) \EqDef \frac{3\pi}{2R}
    - \frac{3\AreaC(\Ball{R/h}{\vx/h} \cap
    {Z}, h)}{R^3}\,,\label{eq:dig-curvature-estimator-k}\\
    &\MeanCurvH{R}(Z,\vx,h) \EqDef \frac{8}{3R}
    - \frac{4\VolC(\Ball{R/h}{\vx/h} \cap
    Z, h)}{\pi R^4}\,.\label{eq:dig-curvature-estimator-H}
  \end{align}
  %
  where $\Ball{R/h}{\vx/h}$ is the ball of digital radius $R/h$ centered on
  digital point $(\vx/h) \in Z$.
  %
  % \AreaC(\Ball{R/h}{\vx/h} \cap \Body{\DSh}{h}, h) \EqDef h^2 \MCard(
  % \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ and $\VolC(\Ball{R/h}{\vx/h} \cap
  % \Body{\DSh}{h}, h) \EqDef h^3 \MCard( \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ are
  % the number of digital points in the intersection between $\Body{\DSh}{h}$ and
  % the ball $\Ball{R/h}{\vx/h}$ of radius $R/h$ centered on $\vx/h$, scaled by
  % $h^d$.\todo[inline]{J: Hum... pas top...}
  %
  % Similarly, $\hat{\lambda}_1$
  % and $\hat{\lambda}_2$ are the two greatest eigenvalues of the covariance
  % matrix of the digital points in the intersection between $Z$ and the ball.
  %
\end{Definition}
%
They have proved that estimating both 2D curvature and 3D mean curvature on digital
objects with volumetric integral information have the multigrid convergence
property \cite{DGCI2013}. For short, when the digital object becomes finer and
finer, \ie when the digitization step $h$ tends to zero, the estimated quantity
on $\Bd{\Body{\DSh}{h}}$ converges (theoretically and experimentally) to the
associated one on $\Bd{\Shape}$.
%
%Since our digital estimators are defined on
%digital shapes, in order to compare results with the exact curvature on
%Euclidean shape, we need to match points on boundary of the digital
%shape and points on the border of the Euclidean one.
%
Setting $R = kh^\frac{1}{3}$ on convex shapes
with at least $C^3$-boundary and bounded curvature, where $k$ is a constant
related to the maximal curvature of the shape, they have:
%
\begin{align}
\forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
\forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
&\CurvH{R}(\DSh, \vxH, h) = \Curv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,,\\
&\MeanCurvH{R}(\DSh, \vxH, h) = \MeanCurv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,.
\end{align}

@@

Additionally, eigenvectors associated to $\hat{\lambda}_1$ and
$\hat{\lambda}_2$ of the covariance matrix are principal
curvature direction estimators $\PrincDirH{1}{R}$ and
$\PrincDirH{2}{R}$. The smallest eigenvector corresponds to the normal
direction $\NormalDirH{R}$ at $\vx$. Convergence results can be found
in \cite{ChapterIICurvature}.
%%% Changer avec le chapitre de Jaco à un moment

@@

Representing a hierarchical structure on GPU is usually challenging since
such a data parallel component is unable to handle recursivity.

@@

Similarly to original MC algorithm, this approach is well suited %perfectly adapted
to a GPU implementation: given a set of cells (a vector of morton codes),
each triangle patch can be extracted in parallel for both regular cells and transition cells.
%(done by the \emph{geometry shader} in the GPU graphic pipeline).

@@

As illustrated in Figure~\ref{fig_quadtree_partitionning}-\emph{middle}, we
%propose a viewpoint dependent criterion to decide if a cell needs to be refined:
propose a viewpoint dependent criterion to decide if a cell needs to be refined:
the closer we are to the camera, the finer the cells are.
%In addition to the
%distance criterion, we define an angular criterion to prevent refining cells
%outside the camera frustrum.
Such a criterion is also well suited to GPU since it can be evaluated
independently on every cell.
%Hence, we can efficiently determine if a cell must be subdivided or not.
Figure~\ref{fig_lod_octree} illustrates our level of
details (LoD for short) criterion.
% we use. More precisely,
In dimension 2, if $\alpha$ denotes the viewing angle, an object at a distance $d$ from the camera
%with diameter greater than $2\cdot d\cdot\tan(\alpha)$ is projected into more
has a projected size on screen of $2\cdot d\cdot\tan(\alpha)$.
%than one pixel
(see Figure~\ref{fig_lod_octree}-\emph{left}).
%Hence, the
Our distance criterion is based on the ratio (\emph{visibility ratio} in the
following) between the cell diameter $l(c)$ (power of 2 depending on the depth),
and its projected size. %$2\cdot d_c\cdot\tan(\alpha)$, $d_c$ being the distance between the cell
%center and the camera.
For a given cell $c$, split and merge decision are based
on this visibility ratio:
%
\begin{itemize}
  \item $c$ is split if its children
cells have a visibility ratio greater than  constant $k$\,;
  \item $c$ and its sibling cells are merged if their parent cell $c'$ has a visibility ratio lower than $k$\;
  \item otherwise, the cell $c$ stays for the next frame.
\end{itemize}
%
\begin{figure}[!h]
  \begin{center}
    \begin{overpic}[width=3.5cm]{figs/criterion}
      \put(33,40){$\alpha$}
      \put(50,45){$d_c$}
      \put(95,65){$l(c)$}
    \end{overpic}~~~~~
    \includegraphics[width=3cm]{viewlod2_small}~~~~~
    %\includegraphics[width=3cm]{subdivision}
  \end{center}
  \caption{Notations \emph{(left)} and adaptive meshing in dimension 2 using the LoD distance and angular criterion \emph{(right)}.
  %Based on LoD criterion evaluation, the cell buffer is updated between each frame \emph{(right)}.
  }
  \label{fig_lod_octree}
\end{figure}
%
%In addition to the distance criterion, we add an angular term in order
%to prevent subdivision of cells far from the camera frustrum (see
%Figure~\ref{fig_lod_octree}-\emph{middle}). We do not go into the
%details but please note that stay, split and merge decisions of all cells are computed in
%parallel from all cell morton codes.
Using such a criterion, split and merge decisions are computed in
%of all cells are computed in
parallel from the morton codes of all cells.

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{figs/pipeline2}
  \caption{GPU pipeline summary. Data buffers are
    represented in green and computations in red. Each computation
    retrieves data from a buffer and fills a new one. }
  \label{fig_pipeline}
\end{figure}

@@

The general idea is to perform an Integral Invariant computation on
GPU at each vertex of the generated triangulated mesh. First note that
when the LoD criterion is removed, each MC vertex is exactly centered at
a surfel center. At different depth of the octree, MC vertices still
correspond to surfel centers of subsampled versions of the input
object. Hence, Integral Invariant framework defined in
Section~\ref{sec:curv-tens-estim} is consistent with the triangulated
surface obtained on GPU: triangles are used for visualization purposes
but all computations are performed on the digital object $\DSh$ for
estimators
(\ref{eq:dig-curvature-estimator-k}),
(\ref{eq:dig-curvature-estimator-H}), (\ref{eq:dig-curvature-estimator-k1})
and (\ref{eq:dig-curvature-estimator-k2}).

To implement the integration on $\Ball{R}{\vx} \cap \Shape$
(Fig.~\ref{fig:approx}-$a$), a naive approach consists in scanning all digital
points inside such domain (Fig.~\ref{fig:approx}-$b$) and then to estimate the
geometrical moment as the sum of the geometrical moments of each elementary
cubes lying in the intersection. On GPU, \emph{mipmap} textures are used to
obtain a multi-resolution information: if the input binary object is stored in a
3D texture, GPU hardware can construct a multi-resolution pyramid such that 8
neighboring voxel intensities at level $l$ are averaged to define the voxel
value at level $l+1$. If the level 0 corresponds to the binary input object, at
a given level $l$, a voxel $(x,y,z)$ contains the fraction of $\DSh$ belonging
to the cube of center $(x,y,z)$ and edge length $2^l$. As a consequence, we can
approximate the volume of $\Ball{R}{\vx}\cap\Shape$ by considering mipmap values
at a given resolution $l$ (Fig.~\ref{fig:approx}-$c$). In this case, errors only
occurs for cells lying inside $\Shape$ (with density 1) not entirely covered by
$\Ball{R}{\vx}$. Furthermore, the \emph{mipmap} texture can be used to design,
using a single texture probe, a fast inclusion test of a given cell $c$ at level
$l$ into the shapes: we say that $c$ is in $\Shape$ if its density (retrieved
from the texture probe at level l) is greater than $1/2$. The idea here is to
mimic a kind of adaptive Gauss digitization process.

@@

Using the hierarchical nature of the mipmap texture, we could also
consider hierarchical decompositions of either $\Ball{R}{\vx}$
(Fig.~\ref{fig:approx2}-$a$) or $\Ball{R}{\vx}\cap\Shape$
(Fig.~\ref{fig:approx2}-$b$). The idea would be to decompose the
regions into mipmap cells of different resolution in order to limit
the number of texture access (important bottleneck on GPU hardwares)
and to get better approximated quantities. Note that the scenario in
Fig.~\ref{fig:approx2}-$b$ is clearly the best approach: the
volume is exact  (\wrt the regular grid case) since geometrical moments
are additive quantities, and the number of texture fetches is  optimal
at this point.

@@

\paragraph{Hierarchical Decomposition.}  As discussed above, best
results in terms of texture fetch can be obtained by the approach
presented in Fig.~\ref{fig:approx2}-$b$. At each vertex, the
computation algorithm would consist in a recursive decomposition of
the domain such that a cell is subdivided if it does not belong to
$\Ball{R}{\vx}\cap\Shape$. However, on the GPU %hardware,
this approach is
highly inefficient since the hardware optimizes the parallelism only
when the micro-program described in the shader has a predictive
execution flow. Hence, implementing the recursive algorithm (or its
de-recursified version) requires a lot of branches (conditional
\emph{if} tests) in the flow. Instead of implementing this approach,
we have considered a fast hierarchical approximation following
Fig.~\ref{fig:approx2}-$a$: for a given radius $R$, we precompute a
hierarchical decomposition of $B_R$. Such decomposition is made of
octree cells at different resolutions. At a given MC vertex $\vx$, the
algorithm becomes simple since we just scan the ball cells and
estimate the area (or other moments) from the mipmap values associated
to this cell. Note that cells in the $B_R$ decomposition may not be
aligned with mipmap cells. However, we can use GPU %mipmap
functionalities which interpolates mipmap values if we probe at
non-discrete positions. For a given radius $R$, the expected number of
cells in $B_R$ is in $O(\log{R})$. Implementation details on the
hierarchical octree representation can be found in the supplementary
material (see Section \ref{sec:discussion}).
